---
title: "Assignment"
author: "Xiaohan Xie"
format: html
editor: visual
---

### 1. Load packages
```{r}
suppressPackageStartupMessages({
  library(AnnotationHub) # to fetch genomes, annotations, etc.
  library(ensembldb)     # to manipulate ensembl gene annotations
  library(GenomicRanges) # data structure for genomic ranges 
  library(epiwraps)      # visualization
})
```
### 2. Find and download following annotations data
2.1 The mouse (Mus Musculus) EnsDb object, version 102, genome build GRCm38
```{r}
ah <- AnnotationHub()
query_results <- query(ah, c("Mus Musculus", "EnsDb"))

query_results$title #[16] "Ensembl 102 EnsDb for Mus musculus"  

query_results$ah_id[16]

ensdb_mouse <- ah[["AH89211"]]
ensdb_mouse
```
2.2 The mouse genome sequence ( dna_sm ) in TwoBit/2bit format for GRCm38
```{r}
mouse_dna_query <- query(ah, c("Mus musculus", "TwoBit", "GRCm38", "dna_sm"))
mouse_dna_query$title
mouse_dna_query$ah_id
rev(mouse_dna_query$ah_id)[1]
dna_mouse <- ah["AH88477"]
dna_mouse
```
2.3 The drosophila melanogaster genome sequence ( dna_sm ) in TwoBit/2bit format for BDGP6
```{r}
fly_dna_query <- query(ah, c("Drosophila melanogaster", "TwoBit", "BDGP6", "dna_sm"))
fly_dna_query$title
rev(fly_dna_query$ah_id)[1]
dna_fly <- ah["AH106160"]
dna_fly
```
### 3. Using the mouse EnsDb, find the following:
3.1 How many different ensembl gene IDs and gene symbols are there for protein-coding genes?
```{r}
mouse_genes <- genes(ensdb_mouse)
mouse_pcg <- mouse_genes[mouse_genes$gene_biotype=="protein_coding"]
num_gene_id <- length(unique(mouse_pcg$gene_id))
num_gene_id
num_gene_symbol <- length(unique(na.omit(mouse_pcg$symbol)))
num_gene_symbol
```
3.2 Plot the distribution of the (spliced) length of protein-coding transcripts
```{r}
filtered_exons <- exonsBy(ensdb_mouse, by = c("tx"), filter = ~ gene_biotype == "protein_coding")
spliced_lengths <- sum(width(filtered_exons))
hist(log10(spliced_lengths), 
     breaks = 100, 
     col = "olivedrab", 
     main = "Distribution of Spliced Lengths (Protein-Coding)",
     xlab = expression(log[10] * " Transcript Length"),
     ylab = "Frequency")
```

