---
title: "assignment07"
author: "Xiaohan Xie"
format: html
editor: visual
self-contained: true
---
## Part 1
In the same dataset of ATAC on chr19 used in the first part of the practical, plot
1) the insertion (i.e. ‘cuts’) profile of nucleosome-free fragments and
2) the centers of nucleosome-containing fragments, around the high-confidence motifs of a transcription factor of your choice (one figure containing the two signals)
Transcription factor chosen: REST

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(AnnotationHub)
  library(MotifDb)
  library(memes)
  library(universalmotif)
  library(ensembldb)
  library(ggplot2)
  library(sechm)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(limma)
  library(betterChromVAR)
})

ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
```

### Obtaining the sites with a REST motif

```{r}
# get CTCF motif
motif <- MotifDb::query(MotifDb, c("REST","Mus"))[[1]]
motif2 <- convert_motifs(motif, class="TFBSTools-PFMatrix")
genome <- ah[["AH68356"]]
# get the sequence for chr19:
chr19 <- import(genome)["19"]

# find motif matches across chr19
moi <- motifmatchr::matchMotifs(motif2, chr19, out="positions", p.cutoff=1e-5)[[1]]
# convert to GRanges
moi <- as(setNames(moi,names(chr19)), "GRanges")
```
### Plotting signal around the motif occurences

```{r, fig.width=8, fig.height=4}
# we prepare the list of tracks
tracks <- list.files(pattern="bw$")
names(tracks) <- gsub("\\.bw","",basename(tracks))

# extract signals around the motif occurences
# we zoom in to 300bp around the motif centers, in windows of 5bp
sm <- signal2Matrix(tracks, moi, w=5, extend=300)

# plot the signals:
plotEnrichedHeatmaps(sm, trim=0.95, multiScale = TRUE, axis_name = "0", use_raster = FALSE)
```
```{r}
d <- meltSignals(sm[,c("NF_cuts", "mono_centers")])
ggplot(d, aes(position, mean, colour=sample)) + geom_line(size=1.2)
```

## Part 2
In the peak counts dataset from stressed brains that we used in the second part of the practical, identify motifs that change in accessibility between males and females, and plot a heatmap of the deviations for the top few candidates motifs.
```{r}
se <- readRDS("peakCounts.rds")
```
### Annotate regions
```{r}
# we restrict the regions to standard chromosomes and change the seq level style to match the genome
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- addGCBias(se, genome=BSgenome.Mmusculus.UCSC.mm10)
# (to use a fasta file as genome, we could simply do genome=FaFile("path/to/genome.fasta") )
```
### Motif matches
```{r, message=FALSE}
# make sure you specify a collection, in order to avoid having redundant motifs:
motifs <- MotifDb::query(MotifDb, "HOCOMOCOv13")
motifs <- do.call(TFBSTools::PWMatrixList, convert_motifs(motifs, class="TFBSTools-PWMatrix"))
names(motifs) <- make.unique(sapply(motifs, \(x) gsub("\\..+","",x@name)))

matches <- motifmatchr::matchMotifs(motifs, se, genome=BSgenome.Mmusculus.UCSC.mm10)
```
### Get Motif deviations
```{r}
dev <- betterChromVAR(se, matches)
```
### Differential Analysis between Males and Females
```{r}
dev$group <- ifelse(grepl("female", colnames(dev)), "Female", "Male")
design <- model.matrix(~group, data=as.data.frame(colData(dev)))
fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, number = Inf))
res$motif <- row.names(res)
ggplot(res, aes(logFC, -log10(adj.P.Val), label=motif)) + geom_text()
```
```{r}
sechm(dev, head(row.names(res)), assayName = "z")
```
