---
title: "assignment06"
author: "Xiaohan Xie"
format: html
self-contained: true
editor: visual
---

```{r setup}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(GenomicRanges)
  library(ggplot2)
  #library(memes) # for the meme-based methods(still need to install MEME) -- COMMENT OUT when using alternatives 
  library(motifmatchr) # for scanning sequences for matches of given motifs
  library(Biostrings) # for handling sequences
  library(MotifDb) # database of motifs
  library(TFBSTools) # for handling some motif formats
  library(universalmotif) # for converting motifs to various formats
  library(PWMEnrich) # for R-based motif enrichment analysis
  library(GenomeInfoDb)
  library(ensembldb)
  library(rtracklayer) 
  library(dplyr)
})
```

## Obtaining peak sequences

Transcription factor chosen: EGR1 Biosample chosen: Homo sapiens liver tissue female child (4 years)

```{r}
download.file("https://www.encodeproject.org/files/ENCFF768POQ/@@download/ENCFF768POQ.bed.gz", "hsLiver_EGR1_ENCFF768POQ.bed.gz")
peaks <- rtracklayer::import("hsLiver_EGR1_ENCFF768POQ.bed.gz", format="NarrowPeak")
seqlevelsStyle(peaks) <- "Ensembl" 
head(peaks)
```

Get the corresponding GRCh38 genome:

```{r}
ah <- AnnotationHub()
query_results <- AnnotationHub::query(ah, c("Homo Sapiens", "TwoBit", "GRCh38"))
query_df <- data.frame(query_results$ah_id, query_results$rdatadateadded) 
query_df %>% arrange(desc(query_results.rdatadateadded))
genome <- ah[["AH82402"]]
genome_seqs <- import(genome)
peak_seqs <- Biostrings::getSeq(genome, peaks)
```

## Identify instances of the factor's motif

### Get the desired motif

```{r}
# we search for "EGR1" in the motif database
motifs <- query(MotifDb, c("EGR1"))
# there are several matching motifs:
names(motifs)
# we select one:
motif <- motifs[["Hsapiens-HOCOMOCOv10-EGR1_HUMAN.H10MO.S"]]
motif
# we visualize it:
#view_motifs(motifs[1:2])
view_motifs(motif)
```

### Using motifmatchr:

```{r}
# convert the sequences to fasta:
Biostrings::writeXStringSet(genome_seqs, "genome_seqs.fa")
motif2 <- convert_motifs(motif, class="TFBSTools-PWMatrix")
motif_hits <- motifmatchr::matchMotifs(motif2, subject=peaks, genome=Rsamtools::FaFile("genome_seqs.fa"),
                                out="positions")
# the higher the score the better
motif_hits <- motif_hits[[1]] # we scanned for just one motif, so we get the results for that motif
head(motif_hits)
hist(motif_hits$score)
```

## Of all the peaks, what proportion contains a motif for the factor?

```{r}
overlaps <- overlapsAny(peaks,motif_hits) # cannot use within here as we need the proportion of peaks having overlaps
counts_contain <- sum(overlaps)
proportion_contain <- mean(overlaps)
sprintf("Of the %d peaks, %d (%.1f%%) contain a motif.", 
        length(peaks), counts_contain, proportion_contain*100)
```

## Of all instances of that motif in the genome, what proportion is bound by the factor (i.e.has a peak)?

```{r}
motif_across_genome <- matchMotifs(motif2, subject=genome_seqs, out="positions")[[1]]
names(motif_across_genome) <- names(genome_seqs)
motif_across_genome <- as(motif_across_genome, "GRanges")
head(motif_across_genome)
```

```{r}
overlaps <- overlapsAny(motif_across_genome, peaks) # could use within to be stricter ?
counts_contain <- sum(overlaps)
proportion_contain <- mean(overlaps)
sprintf("Of the %d motif instances, %d (%.1f%%) overlap a peak.", 
        length(motif_across_genome), counts_contain, proportion_contain*100)
```

Possible reason for the discrepancy of 7707 >> 2573: when identifying motifs in the peaks sequences using motif_hits <- motifmatchr::matchMotifs(motif2, subject=peaks, genome=Rsamtools::FaFile("genome_seqs.fa"),
                                out="positions"), background is calculated only with peak sequences, many weaker motif occurrences are thus not able to pass the significance threshold. If we specify bg="genome", the overlapping results would be much more similar. 
